Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Region index minimal stuff
Closes #222

To demonstrate the computational accessibility of Zarr on this
large human dataset, we performed some illustrative benchmarks.
As these benchmarks take some time to run, we focus
on a single 132 GiB compressed VCF file covering
positions 58,219,159--60,650,943 (562,640 variants)
from the middle of the list of 106 files for chromosome 2.
We report both the total CPU time and elapsed wall-clock time here
as both are relevant.
First, we extracted the genome position for each variant in this single VCF
chunk using \texttt{bcftools query} and Python Zarr code as described in
Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required
% def parse_time(time_str):
%     minsplit = time_str.split("m")
%     return datetime.timedelta(minutes=int(minsplit[0]),
%         seconds=float(minsplit[1][:-1]))
% def parse_time_output(s):
%     values = {}
%     for line in s.splitlines():
%         split = line.split()
%         values[split[0]] = parse_time(split[1])
%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
% real  85m51.013s
% user  54m39.986s
% sys 0m45.277s
55.42 min CPU and 85.85 min elapsed.
% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
% Wall time: 1min 44s
% 55.42 * 60 / 2.78  = 1196
% 85.85 / 1.73 = 49.62
The Zarr code required 2.78 sec CPU and 1.73 min elapsed.
This is a 1196X smaller CPU burden and a 50X speed-up in elapsed time.
The major difference between CPU time and wall-time is noteworthy
here, and indicates some opportunities for improvement in VCF Zarr
in high-latency environments such as the shared file system in the
Genomics England HPC system. Currently VCF Zarr does not store any
specialised index to map genomic coordinates to array positions
along the variants dimension. Instead, to find the relevant slice
of records corresponding to the range of positions in the target
VCF file, we load the entire variant\_position array and
binary search. This entails reading 5,989 chunk files
(the chunk size is 100,000 variants) which incurs a substantial
latency penalty on this system. Later versions of the specification
may solve this problem by storing an array of size
(approximately) the number variant chunks
which maps ranges of genome coordinates to chunk indexes,
or a more specialised structure that supports overlap queries.

>>>>

To demonstrate the computational accessibility of Zarr on this
large human dataset, we performed some illustrative benchmarks.
As these benchmarks take some time to run, we focus
on a single 132 GiB compressed VCF file covering
positions 58,219,159--60,650,943 (562,640 variants)
from the middle of the list of 106 files for chromosome 2.
We report both the total CPU time and elapsed wall-clock time here
as both are relevant.
First, we extracted the genome position for each variant in this single VCF
chunk using \texttt{bcftools query} and Python Zarr code as described in
Fig~\ref{fig-column-extract}. The \texttt{bcftools} command required
% def parse_time(time_str):
%     minsplit = time_str.split("m")
%     return datetime.timedelta(minutes=int(minsplit[0]),
%         seconds=float(minsplit[1][:-1]))
% def parse_time_output(s):
%     values = {}
%     for line in s.splitlines():
%         split = line.split()
%         values[split[0]] = parse_time(split[1])
%     return values["real"].seconds / 60, (values["user"] + values["sys"]).seconds / 60
% real  85m51.013s
% user  54m39.986s
% sys 0m45.277s
55.42 min CPU and 85.85 min elapsed.
% CPU times: user 846 ms, sys: 1.93 s, total: 2.78 s
% Wall time: 1min 44s
% 55.42 * 60 / 2.78  = 1196
% 85.85 / 1.73 = 49.62
The Zarr code required 2.78 sec CPU and 1.73 min elapsed.
This is a 1196X smaller CPU burden and a 50X speed-up in elapsed time.
The major difference between CPU time and wall-time is noteworthy
here, and highlights the importance of using an index to find
the location of variant chunks.
This benchmark predates the \texttt{region\_index}
array computed by \texttt{vcf2zarr} (see Methods) which 
stores sufficient information to identify which variant chunks intersect
with a given range query.
Here, to find the relevant slice
of records corresponding to the range of positions in the target
VCF file, we simply load the entire \texttt{variant\_position} array and
binary search. This entails reading 5,989 chunk files
(the chunk size is 100,000 variants) and incurred a substantial
latency penalty on the Genomics England HPC shared filesystem.
